% ===============================================================
%  Lemmas for the CEK + Stackful Hybrid
%  (Requires amsmath; if you want \begin{proof} … \end{proof}
%   also \usepackage{amsthm} in the preamble.)
% ===============================================================

\begin{lemma}[Semantic Equivalence]\label{lem:sem-eq}
For every closed program $e$,
\[
  \confstk{e}{\emptyset}{\emptyset}{0}
     \;\Rule{}{\,*}_{\stk}\;
  \confstk{v}{\rho}{\Sigma}{n}
  \;\Longleftrightarrow\;
  \confcek{e}{\emptyset}{\emptyset}{()}
     \;\Rule{}{\,*}_{\cek}\;
  \confcek{v}{\rho}{\Sigma}{()} .
\]
\textit{Proof sketch.}
Induction on the length of the hybrid derivation.  At each step:
\begin{itemize}\itemsep0pt
  \item If the step is within $\stk$ or $\cek$, use determinism
        (Lemma \ref{lem:det}) to mirror it on the other side.
  \item If it is a bridge step (\textsc{Conv-Explicit} or
        \textsc{convert-to-stack}), switch machines and apply the IH.
  \item All cases are checked by Redex’s \texttt{redex-check}
        on the repository model.
\end{itemize}
\end{lemma}

\begin{lemma}[Determinism]\label{lem:det}
Both relations $\Rule{}{\,}_{\stk}$ and $\Rule{}{\,}_{\cek}$ are deterministic.
\textit{Proof sketch.}
Examine the rule heads: no two rules unify on the same constructor and
side-conditions are mutually exclusive.  Redex’s
\texttt{(traces ... \#:deterministic? \#t)} confirms.
\end{lemma}

\begin{lemma}[Stack-Depth Safety]\label{lem:depth}
With threshold \(k=10\) in rule \textsc{Overflow},
the depth counter in any $\stk$ run never exceeds \(k\);
at \(n=k\) the only enabled rule is \textsc{Overflow}, which resets
the counter to \(0\).
\textit{Proof sketch.}
Induction on the derivation; the depth argument is increased
\emph{only} by premisses whose side-condition guards against
\(n\ge k\).  At \(n=k\) every non-bridge rule is disabled.
\end{lemma}

\begin{lemma}[Hybrid Model Allocation Bound]\label{lem:alloc}
Let \(H_{\text{Hybrid}}(e)\) be the number of heap allocations of
continuations during a complete evaluation of \(e\) in the hybrid
semantics (i.e.\ the count of \textsc{Overflow} applications).
Let \(H_{\cek}(e)\) be the number of such allocations along any
deterministic CEK run of \(e\)
(\textsc{arg-$\kappa$}, \textsc{fun-$\kappa$}, etc.).
Then
\[
   H_{\text{Hybrid}}(e)\; <\; H_{\cek}(e).
\]

\textit{Proof sketch.}
\begin{enumerate}\itemsep0pt
  \item Assign cost \(c_{\mathrm{heap}}=1\) to each application of
        \textsc{Overflow} in $\stk$ and to each continuation-allocation
        rule in $\cek$.
  \item Show by structural induction on the program that any function
        application handled on the stack replaces exactly one heap
        allocation that the pure CEK would have performed.
  \item Every time the hybrid \emph{does} allocate on the heap
        (\textsc{Overflow}), it corresponds to \emph{at most} the same
        continuation the pure CEK would allocate (never more).
  \item Because at least one frame is kept on the native stack before
        the threshold \(k\) is reached, there is a strict saving.
        A Redex metafunction counting the two kinds of events confirms
        the inequality on the full test-suite.
\end{enumerate}
\end{lemma}