\begin{centering}
\textbf{ACKNOWLEDGEMENTS}\\
\vspace{\baselineskip}
\end{centering}

\paragraph{}%
  No research is ever conducted, and no dissertation is ever written, by a single person; over the years, everyone who touched my life contributed a piece to the work we built together.

\paragraph{}%
  My journey at IU Computer Science began in Lindley Hall 101—a workshop where generations of PL researchers trained by plunging into the masters' discussions and getting their hands dirty from day one. While the physical surroundings changed over the years, the same spirit I saw in Lindley Hall lived on in PL Wonks. I'm grateful to the PL Wonks regulars for their attention, suggestions, and good times over the years; they will always have a special place in my heart: Deyaaeldeen Almahallawi, Ömer Sinan Ağacan, Spenser Bauman, Ambrose Bonnaire-Sergeant, Laurel Carter, Buddhika Chamith, Tianyu Chen, Vikraman Choudhury, David Christiansen, Matteo Cimini, Dan Friedman, Fred Fu, Matthew Heimerdinger, Jason Hemann, Aaron Hsu, Andrew Kent, Chaitanya Koparkar, Paulette Koronkevich, Andre Kuhlenschmidt, Weixi Ma, Praveen Narayanan, Ryan Newton, Amr Sabry, Ryan Scott, Jeremy Siek, Sarah Spall, Cameron Swords, Sam Tobin-Hochstadt, Mike Vitousek, Michael Vollmer, Victoria Vollmer, Chris Wailes, and Rajan Walia.

\paragraph{}%
  Support from the Racket community—especially Matthew Flatt, Matthias Felleisen, Robby Findler, John Clements, and Shriram Krishnamurthi—brought great insights and delightful discussions at Racket meetings; above all, my advisor, Sam Tobin-Hochstadt, provided steady guidance and patience over the years—even when I was reluctant to use my brain more than absolutely necessary. The PyPy community, especially CF Bolz-Tereick, offered support and guidance as I learned how meta-tracing JITs work. I am grateful to my committee—Sam Tobin-Hochstadt, Jeremy Siek, Amr Sabry, and Daniel Leivant—for their support and guidance. Finally, special thanks go to my first mentor in computer science, Chris Stephenson, who taught me to program and sparked my passion for programming language studies.

\paragraph{}%
  Last, and certainly not least, I thank my family, who sacrificed so much in so many ways to make this happen: Muharrem Derici, Ayşe Derici, Özlem Derici Şengül, Rüştü Derici; Tolga, Semra, Mert, Batu, and all the Gül family; as well as Craig \& Kathy Taylor, Nick \& Amanda Hilton, Gage, and Harlow, all of whom took care of me and became my second family during this research long before I became their son- and brother-in-law. Finally, my incredible wife, Dr. Jennifer Taylor Derici, was not only my greatest support in my PhD journey but also became central to my life during it. Together with our son, Dean, she supported me in countless ways and made this study possible.

\paragraph{}%
  Every significant body of work stands on the shoulders of a generous technical and social community; I'm deeply grateful to this group of people—I couldn't have asked for better.


