\chapter*{Vita}
\addtocontents{toc}{
 \unexpanded{\unexpanded{\renewcommand{\cftchapdotsep}{\cftnodots}}}%
}
\addcontentsline{toc}{chapter}{Curriculum Vitae}

\doublespacing

% ---------- Vita (self-contained) ----------
% Minimal helpers (no extra packages required)
\newlength{\cvdatewidth}\setlength{\cvdatewidth}{3.2cm}
\newcommand{\vitaSection}[1]{\par\vspace{0.75em}\noindent{\Large\textsc{#1}}\par\vspace{0.25em}\hrule\vspace{0.6em}}
% Smart \cvitem: if the left arg is empty, use full width (fixes short wrapping in Publications)
\makeatletter
\newcommand{\cvitem}[2]{%
  \noindent
  \if\relax\detokenize{#1}\relax
    \parbox[t]{\linewidth}{#2}\par\vspace{0.35em}% full width
  \else
    \parbox[t]{\cvdatewidth}{\raggedright #1}\hspace{1em}% date col
    \parbox[t]{\dimexpr\linewidth-\cvdatewidth-1em\relax}{#2}\par\vspace{0.35em}% body col
  \fi}
\makeatother
% Tight linebreak helper to reduce the double-spaced gap between two consecutive lines
\newcommand{\tightline}{\\[-0.35\baselineskip]} % tweak -0.35 as needed

% Header
\begin{center}
{\Large \textsc{Caner Derici}}\\
\small caner@dericilab.live
\end{center}

% Education
\vitaSection{Education}

\cvitem{Sep 2025}{\textbf{PhD in Computer Science}, \textsc{Indiana University}, Bloomington, IN\\Dissertation: \smartQL \thesisTitle \smartQR}
\cvitem{2013}{\textbf{M.S. in Computer Science}, \textsc{Boğaziçi University}, Istanbul, Turkey\\Areas: Machine Learning, Natural Language Processing}
\cvitem{2010}{\textbf{B.S. in Computer Science}, \textsc{Istanbul Bilgi University}, Istanbul, Turkey}

% Work (positions / appointments)
\vitaSection{Work}

\cvitem{2021--2024}{\textbf{Software Engineer II (L4), Enterprise Cloud Engineering, Juju team}\tightline \textsc{Canonical USA}, Remote, US\\[-0.25\baselineskip]
\begin{itemize}\itemsep0.25em
    \vspace{-0.5cm}
  \item Architected distributed components improving reliability, fault tolerance, and back-pressure on an eventually consistent backend.
  \item Doubled the user base of client libraries (Python-libjuju, Terraform Juju Provider); owned deliverables and maintained release cadence.
  \item Contributed to roadmap planning, cross-team coordination, mentoring, and hiring.
\end{itemize}
}

\cvitem{2015--2021}{\textbf{Course Instructor, Teaching \& Research Assistant}\tightline \textsc{Indiana University}, Bloomington, IN\\[-0.25\baselineskip]
\begin{itemize}\itemsep0.25em
    \vspace{-0.5cm}
  \item Taught data structures \& algorithms, compilers, virtual machines, and domain-specific languages.
  \item Researched runtime performance of \gls{jit}-compiled virtual machines for functional languages.
\end{itemize}
}

\cvitem{2012--2013}{\textbf{Software Engineer}\tightline \textsc{Asseco SEE Group}\\[-0.25\baselineskip]
\begin{itemize}\itemsep0.25em
    \vspace{-0.5cm}
  \item Developed virtual payment platforms for e-commerce; delivered three point-of-sale projects in one year (Java, Apache Tomcat, Spring, Mercurial, Jira).
\end{itemize}
}

% Publications (selected)
\vitaSection{Publications}

\cvitem{}{\textit{Flatt, M.}, \textit{Derici, C.}, Dybvig, R.\,K., Keep, A.\,W., et al. “Rebuilding Racket on Chez Scheme (experience report)”. \textit{ICFP} 2019.}
\cvitem{}{\textit{Derici, C.}, et al. “A closed-domain question answering framework using reliable resources to assist students”. \textit{Natural Language Engineering} 2018.}
\cvitem{}{\textit{Derici, C.}, et al. “Question analysis for a closed domain question answering system”. \textit{CICLING} 2015.}
\cvitem{}{\textit{Derici, C.}, et al. “Rule-based focus extraction in Turkish question answering systems”. \textit{SIU} 2014.}
\cvitem{}{Başar, R.\,E., \textit{Derici, C.}, Şenol, Ç. “World With Web: A compiler from world applications to JavaScript”. Technical Report, \textit{Scheme and Functional Programming Workshop} 2009.}

% ---------- End Vita ----------

\pagenumbering{gobble}
